---
title: "<span style='color:green'>Species Distribution Modeling:</span> <span style='color:blue'>Theory and Practice</span>"
author: 
  - name: "Dr. agr. Wyclife Agumba Oluoch"
    affiliation: "[Land Economics Group](https://www.ilr1.uni-bonn.de/en/research/research-groups/land-economics), Institute for Food and Resource Economics, University of Bonn, Bonn, Germany"
date: today
format: 
  html:
    number-sections: true
    code-line-numbers: true
    code-block-bg: "#e5e5e5"
    fig-cap-location: bottom
    fig-align: center
    toc: true
crossref:               # Required for automatic numbering
  chapters: false       # Unless you're writing a book with chapters
bibliography: references.bib
---

# Introduction

Species distribution modeling (SDM) relates species occurrence data to environmental predictors with the aim of estimating habitat suitability across space.

This document provides a minimal, reproducible SDM workflow in $R$ using the `sdm` package [@sdm] version 1.2.67.

## Loading required library

```{r}
#install.packages('sdm') # In case you do not have it installed already
library(sdm)
```

## Reading in the occurrence data

We will use data which comes with the package as that would be the easiest way to begin using the package. This data has presence (1) and absence (0) records of a virtual species. We will read it into $R$ using the vect function from `terra` package [@abatzoglou2018] which is loaded whenever you load `sdm`.

```{r}
#| eval: true
sp <- vect(system.file("external/species.shp", package = "sdm"))
sp
```

## Reading in predictor variables

Apart from the occurrence records, we also need to load the predictor variables. There are four raster layers which come with the package and we will load them alongside the occurrence records to help in building the model. These are also often referred to as response variables. We will use rast function also from `terra` package [@abatzoglou2018].

```{r}
preds <- rast(list.files(system.file("external", package = "sdm"), pattern = ".asc$", full.names = TRUE))
preds
```

## Visualize the data

It is important to visualize the output so that you confirm how the occurrence records, both presence and absence, are distributed over the environmental variable(s), (see @fig-fig1).

```{r}
#| eval: true
#| label: fig-fig1
#| fig-cap: "Distribution of occurrence points over the region of interest"
plot(preds$vegetation)
plot(sp[sp$Occurrence == 1, ], col = "blue", add = TRUE)
plot(sp[sp$Occurrence == 0, ], col = "red", add = TRUE)
```

## Build the `sdm` data object

Once you are satisfied with the distribution of the occurrence records, the next step is to create an $sdmData$ object. This is taking both the occurrence data and predictor variables.

```{r}
#| eval: true
d <- sdmData(
  formula = Occurrence ~.,
  train = sp,
  predictors = preds
)
d
```

The $d$ object tells us information like number of species in our data set, the name of the species, number of features (predictor variables) and their names, type of the data such as presence-absence or presence-only or presence-background, whether we have separate/independent test data, number of records and whether they have coordinates. This is the data we supply to the `sdm` function to train the model.

## Build the model

In building the `sdm` model, we provide our formula and the $d$ object. We also provide the methods, replication method, fraction of the data for testing, and the number of runs. Here we can also provide other arguments like parallelSetting in case we want to use more cores in the computer and modelSettings to alter the default settings of each method.

```{r}
#| eval: true
m <- sdm(
  formula = Occurrence ~.,
  data = d,
  methods = "rf", # There are many methods, see getMethodNames()
  replication = "sub", # Can also use cv or boot.
  test.percent = 20,
  n = 10
)
m
```

## Check few plots

We can make a few plots including the roc curve plot, to check whether the model performed better than random chance, @fig-fig2.

```{r}
#| eval: true
#| label: fig-fig2
#| fig-cap: "Receiver perating Characteristic Curve"
roc(m)
```

We may also want to assess how our the suitability of habitat varies with the changes in values of the predictor variables. Ideally, we expect to see uni-modal curves if we exhausted the niche of the species.

```{r}
#| eval: true
#| label: fig-fig3
#| fig-cap: "Response Curve for the species to environmental variables"
rcurve(m)
```

To see which of the predictor variables was most important for defining the niche of the species, we can use getVarImp function as follows:

```{r}
#| eval: true
#| label: fig-fig4
#| fig-cap: "Variable importance for the species" 
plot(getVarImp(m))
```

This shows us that vegetation, in this case, was the most important variable in explaining the niche of our species of interest.

## Predict

Finally, the end game is to make prediction of habitat suitability of the species over the geographic space. Here we feed all the predictor variables to the model in order to get a continuous habitat suitability map as follows:

```{r}
#| eval: true
#| label: fig-fig5
#| fig-cap: "Current habitat suitability for the species"
p <- predict(object = m, preds, mean = TRUE)
plot(p)
```

## Binary Maps

After creating the habitat suitability map, we may be interested in actually showing where the species are present or absent. For that we need to binarize the suitability map so that we have presence (1) or absence (0). However, since the suitability map is continuous, it is difficult to strike a balance on where to pick the threshold. For this, we can go with a threshold that maximizes sensitivity and specificity. However, depending on the research objective, one may decide to use different threshold. For example, when trying to identify potential areas for conserving an endangered species, one may use lower thershold so that more area is considered present locations and demarcated for conservation, etc. Here we will go with the threshold that maximizes sensitivity and specificity.

```{r}
#| eval: true
#| label: fig-fig6
#| fig-cap: "Presence absence binary map for the species"
pa <- pa(x = p, y = m, id = "ensemble", opt = 2)
plot(pa, col = c("darkgreen", "gray80"))
```

## How much area is suitable for the species?

We can get this information using the expanse function in terra as follows:

```{r}
crs(pa) <- crs(sp)
present_area <- expanse(pa, unit = "ha", byValue = TRUE)
present_area |> 
  mutate(Class = case_when(value == 1 ~ "Presence",
                           value == 0 ~ "Absence",
                           TRUE ~ as.character(value))) |> 
  ggplot(aes(Class, area, fill = Class)) +
  geom_col() +
  # Add text labels on top of the bars
  geom_text(aes(label = round(area)),  # Display area value, rounded to 2 decimals
            vjust = -0.5,                 # Adjust vertical position (move up)
            color = "black",               # Set text color
            size = 5) +                    # Set text size
  scale_fill_manual(values = c("Absence" = "gray80", "Presence" = "darkgreen")) +
  theme_classic(base_size = 16) +
  theme(legend.position = "none") +
  # Optional: Adjust y-axis limits to make room for the text
  ylim(0, max(present_area$area) * 1.1) # Extends the y-axis limit by 10% to prevent clipping
```

# References {.unnumbered}
